---
title: "Preliminary Analysis - Effect of Economic Crisis"
author: "Sneha Vasanth, Janani Kumar, Tigh Bradley, Gaurav Gada"
date: "February 21, 2017"
output: html_document
---

Github link: https://github.com/snehavcs/Data-Science-I-Project

The financial crisis in 2007-2009 occurred as a result of risky financial dealings in the real estate market which spilled over into the world economy at large. The resulting economic downturn in the USA, particularly focusing on the economic changes before and after the collapse of Lehman Brothers in September-October 2008, provides an opportunity to compare impact  of the crisis on national ecnomonies. Canada, China, India, and the USA experienced different economic changes during this period, including dramatic changes in unemployment rates.

This study explores what key economic indicators (GDP, CPI, and Unemployment rate) might have looked like in each of these countries if there had been no economic crisis in 2007-2009.  Arima forecasts are utilized to project how ecnomic trends may have continued if the crisis had not occurred.  The forecast models for the 'healthy' economic state will be compared to what  actually occured in these econmies, providing a measure of the relative economic advantages and disadvantages each country experienced during this period.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(dplyr)
```

Cleaning up the data!

```{r load_cpi}
cpi = read.csv("CPI_World_Bank.csv",check.names = FALSE)
#Get country and indicator column names and remove spaces using make.names
country_indicator_names <- make.names(names(cpi[c(1:4)])) 
#country_indicator_names

#Get year column names
year_column_names <- names(cpi[c(5:ncol(cpi))])
#year_column_names

#Reassign the modified column names to actual column names
names(cpi) <- c(country_indicator_names,year_column_names)
names(cpi)

#Gather all year column names into a new column called Year and assign its values to a column called CPI
clean_cpi <- gather(cpi,Year,CPI,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
cpi_data <- clean_cpi[,!colnames(clean_cpi) %in% c("Indicator.Name","Indicator.Code")]
head(clean_cpi)


cpi = read.csv("CPI_World_Bank.csv",check.names = FALSE)
#Get country and indicator column names and remove spaces using make.names
country_indicator_names <- make.names(names(cpi[c(1:4)]))
#country_indicator_names

#Get year column names
year_column_names <- names(cpi[c(5:ncol(cpi))])
#year_column_names

#Reassign the modified column names to actual column names
names(cpi) <- c(country_indicator_names,year_column_names)
names(cpi)

#Gather all year column names into a new column called Year and assign its values to a column called CPI
clean_cpi <- gather(cpi,Year,CPI,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
cpi_data <- clean_cpi[,!colnames(clean_cpi) %in% c("Indicator.Name","Indicator.Code")]
head(cpi_data)

unemp = read.csv("Unemployment_ILO.csv",check.names = FALSE)
country_indicator_names <- make.names(names(unemp[c(1:4)])) 
#country_indicator_names

#Get year column names
year_column_names <- names(unemp[c(5:ncol(unemp))])
#year_column_names

#Reassign the modified column names to actual column names
names(unemp) <- c(country_indicator_names,year_column_names)
names(unemp)

#Gather all year column names into a new column called Year and assign its values to a column called unemp
clean_unemp <- gather(unemp,Year,unemp,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
unemp_data <- clean_unemp[,!colnames(clean_unemp) %in% c("Indicator.Name","Indicator.Code")]
head(clean_unemp)

#Get country and indicator column names and remove spaces using make.names
country_indicator_names <- make.names(names(unemp[c(1:4)]))
#country_indicator_names

#Get year column names
year_column_names <- names(unemp[c(5:ncol(unemp))])
#year_column_names

#Reassign the modified column names to actual column names
names(unemp) <- c(country_indicator_names,year_column_names)
names(unemp)

#Gather all year column names into a new column called Year and assign its values to a column called unemp
clean_unemp <- gather(unemp,Year,unemp,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
unemp_data <- clean_unemp[,!colnames(clean_unemp) %in% c("Indicator.Name","Indicator.Code")]
head(unemp_data)


gdp = read.csv("gdp.csv",check.names = FALSE)
country_indicator_names <- make.names(names(gdp[c(1:4)])) 
#country_indicator_names

#Get year column names
year_column_names <- names(gdp[c(5:ncol(gdp))])
#year_column_names

#Reassign the modified column names to actual column names
names(gdp) <- c(country_indicator_names,year_column_names)
names(gdp)

#Gather all year column names into a new column called Year and assign its values to a column called gdp
clean_gdp <- gather(gdp,Year,gdp,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
gdp_data <- clean_gdp[,!colnames(clean_gdp) %in% c("Indicator.Name","Indicator.Code")]
head(clean_gdp)

#Get country and indicator column names and remove spaces using make.names
country_indicator_names <- make.names(names(gdp[c(1:4)]))
#country_indicator_names

#Get year column names
year_column_names <- names(gdp[c(5:ncol(gdp))])
#year_column_names

#Reassign the modified column names to actual column names
names(gdp) <- c(country_indicator_names,year_column_names)
names(gdp)

#Gather all year column names into a new column called Year and assign its values to a column called gdp
clean_gdp <- gather(gdp,Year,gdp,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
gdp_data <- clean_gdp[,!colnames(clean_gdp) %in% c("Indicator.Name","Indicator.Code")]
head(gdp_data)


# Join all data
gdp_cpi = merge(gdp_data, cpi_data, by=c("Country.Code","Year"), all = T) # NA's match

labor_data = merge(gdp_cpi, unemp_data, by=c("Country.Code","Year"), all = T) # NA's match

drops = c("Country.Name.y","Country.Name.x")
labor_data = labor_data[ , !(names(labor_data) %in% drops)]

names(labor_data)

labor_data = labor_data[c("Country.Name", "Country.Code", "Year", "gdp", "unemp", "CPI")]

# Clean up data
labor_data = labor_data[!(is.na(labor_data$unemp) & is.na(labor_data$CPI) & is.na(labor_data$gdp)),]

# Clean up the working space
rm(list = ls()[grep("labor", ls(), invert = T)])

country_codes = read.csv("country-codes.csv")

country_data = labor_data[labor_data$Country.Code %in% country_codes$ISO3166.1.Alpha.3,]

labor_data <- country_data 

attach(labor_data)
```

## Exploratory Analysis

How many unique observations to you have?

The number of unique observations for CPI, GDP and unemployment respectively are:

```{r}
nrow(labor_data[!is.na(CPI),])
nrow(labor_data[!is.na(gdp),])
nrow(labor_data[!is.na(unemp),])
```


What information/features/characteristics do you have for each observation?

We modify the scope of our dataset from 1991 to 2014, since unemployment data is only available for this range.

```{r}
labor_data <- labor_data[Year>=1991 & Year<=2014,]
head(labor_data)
```

What are the min/max/mean/median/sd values for each of these features?

```{r summary_stats_cpi}
#Summary values for CPI
summary(labor_data$CPI,na.rm = TRUE)
sd(labor_data$CPI, na.rm = TRUE)

#Summary values for GDP
summary(labor_data$gdp,na.rm = TRUE)
sd(labor_data$gdp, na.rm = TRUE)

#Summary values for Unemployment
summary(labor_data$unemp,na.rm = TRUE)
sd(labor_data$unemp, na.rm = TRUE)
```

We found the averages for CPI, gdp and unemployment for all the countries.
Now, let's manipulate the dataset to include a column to indicate if the time period for the data is before(1991-2006) or after(2007-2014) the economic crisis.


```{r}
labor_data <- labor_data %>% mutate(Time_Period = ifelse(Year < 2007,"Before Crisis","After Crisis"))
head(labor_data)
```

Converting Time_Period to a factor variable and releveling the labels.

```{r}
labor_data$Time_Period <- as.character(labor_data$Time_Period)
labor_data$Time_Period <- factor(labor_data$Time_Period, levels = c("Before Crisis", "After Crisis"))
levels(labor_data$Time_Period)
```


```{r}
ggplot(labor_data, aes(x=Time_Period, y = unemp))+geom_boxplot(na.rm = T)
```

The boxplot confirms the finding the average unemployment is almost the same before and after crisis.
This is an interesting finding since the 2007 economic crisis led to unemployment. We will try to analyze how the unemployment trends varied post recession by drilling down further.


```{r}
ggplot(labor_data, aes(x=Time_Period, y = CPI))+geom_boxplot(na.rm = T)
```

The average value of CPI has increased increased post the crisis but the variability has got much less over time.

```{r}
ggplot(labor_data, aes(x=Time_Period, y = gdp))+ylim(0,100000000)+geom_boxplot(na.rm = T)
```

The average value of gdp has decreased post the crisis but the variability has got much less over time.

What is the distribution of the core features (show a histogram)?

Below are the histograms of the distribution of the CPI, GDP and Unemployment before and after 2007 economic crisis.

```{r hist_cpi}
ggplot(labor_data, aes(x = CPI)) +
  geom_histogram(fill = "white",
                 color = "black",
                 binwidth = 5,na.rm = T)+facet_grid(.~Time_Period)
```


```{r}

ggplot(labor_data, aes(x = gdp)) +
  geom_histogram(fill = "white",
                 color = "black",na.rm = T)+facet_grid(.~Time_Period)

```


```{r}

ggplot(labor_data, aes(x = unemp)) +
  geom_histogram(fill = "white",
                 color = "black",
                 binwidth = 0.5,na.rm = T)+facet_grid(.~Time_Period)
```

The distribution of CPI was significantly different before and after crisis. The count of the number of countries with higher inflation rates increased after crisis. The distribution for GDP and unemployment are not significantly different before and after crisis. 

Let's understand the effect of CPI,GDP and unemployment further by analyzing trend graph from 1991 to 2014 in USA,Canada,China and India.

Are there obvious trends in the data (over time, across subgroups, etc.), and are the differences statistically significant?

```{r}
cpi_data_1 <- labor_data %>%
  select(Country.Name,Country.Code,Year,CPI) %>%
  filter(Year >= 1991 & Year<= 2014 & (Country.Code =='IND' | Country.Code == 'CAN'| Country.Code == 'CHN' | Country.Code == 'USA')) %>%
  group_by(Year)

ggplot(cpi_data_1,aes(x = Year, y = CPI, col = Country.Code, group = Country.Code)) +geom_line(na.rm = T,lwd = 1)+ theme(axis.text.x=element_text(angle=90, hjust=1))
```
There is a steep CPI increase for India and the CPI of USA and Canada are comparitively more stable post the 2008 recession.

```{r}
unemployment_data_1 <- labor_data %>%
  select(Country.Name,Country.Code,Year,unemp) %>%
  filter(Year >= 1991 & Year<= 2014 & (Country.Code =='IND' | Country.Code == 'CAN' | Country.Code == 'USA'|Country.Code == 'CHN')) %>%
  group_by(Year)

ggplot(unemployment_data_1, aes(x = Year, y = unemp,col = Country.Code, group = Country.Code))+geom_line(na.rm = T,lwd = 1)+ theme(axis.text.x=element_text(angle=90, hjust=1))
```

We see that the unemployment rates increases drastically in Canada and USA after 2007 crisis(2008 recession). 
The unemployment rates in China is stable wheras for India it decreases after the 2007 crisis.This can be due to 
increase in outsourcing.

What are the other salient aspects of the data (e.g. geospatial factors, text content, etc.)

The following plots are forecasts for each of the selected country's future unemployment rates based on the 2000-2009 data.  This provides a projection of what the unemployment rates might have looked like in the absence of the financial crisis.

```{r}
#canada
#install.packages(forecast)
library(forecast)
unemployment_can <- unemployment_data_1[unemployment_data_1$Country.Code=="CAN",]
auto.arima(unemployment_can$unemp)
fit <-arima(x=unemployment_can$unemp[1:9], order=c(0,0,1))
preds <-forecast.Arima(fit, h=12)
plot.forecast(preds, xlab = "Canada Unemployment forecast using 2000-2009 data")
#us
unemployment_us <- unemployment_data_1[unemployment_data_1$Country.Code=="USA",]
auto.arima(unemployment_us$unemp)
fit <-arima(x=unemployment_us$unemp[1:9], order=c(0,1,0))
preds <-forecast.Arima(fit, h=12)
plot.forecast(preds, xlab = "USA Unemployment forecast using 2000-2009 data")
#India
unemployment_ind <- unemployment_data_1[unemployment_data_1$Country.Code=="IND",]
auto.arima(unemployment_ind$unemp)
fit <-arima(x=unemployment_ind$unemp[1:9], order=c(0,1,0))
preds <-forecast.Arima(fit, h=12)
plot.forecast(preds, xlab = "India Unemployment forecast using 2000-2009 data")
#china
unemployment_china <- unemployment_data_1[unemployment_data_1$Country.Code=="CHN",]
auto.arima(unemployment_china$unemp)
fit <-arima(x=unemployment_china$unemp[1:9], order=c(1,0,0))
preds <-forecast.Arima(fit, h=12)
plot.forecast(preds, xlab = "USA Unemployment forecast using 2000-2009 data")

```

Given the different models auto.arima recommended above, a second attempt follows utilizing the unemployment data from 1991-2009:

```{r}
unemployment_data_2 <- country_data %>%
  select(Country.Name,Country.Code,Year,unemp) %>%
  filter((Country.Code =='IND' | Country.Code == 'CAN' | Country.Code == 'USA'|Country.Code == 'CHN')) %>%
  group_by(Year)
#canada
unemployment_can <- unemployment_data_2[unemployment_data_2$Country.Code=="CAN",]
auto.arima(unemployment_can$unemp[32:50])
fit <-arima(x=unemployment_can$unemp[32:50], order=c(0,1,1))
preds <-forecast.Arima(fit, h=12)
plot.forecast(preds, xlab = "Canada Unemployment forecast using 1991-2009 data")
#us
unemployment_us <- unemployment_data_2[unemployment_data_2$Country.Code=="USA",]
auto.arima(unemployment_us$unemp[32:50])
fit <-arima(x=unemployment_us$unemp[32:50], order=c(2,0,1))
preds <-forecast.Arima(fit, h=12)
plot.forecast(preds, xlab = "USA Unemployment forecast using 1991-2009 data")
#India
unemployment_ind <- unemployment_data_2[unemployment_data_2$Country.Code=="IND",]
auto.arima(unemployment_ind$unemp[32:50])
fit <-arima(x=unemployment_ind$unemp[32:50], order=c(0,0,0))
preds <-forecast.Arima(fit, h=12)
plot.forecast(preds, xlab = "India Unemployment forecast using 1991-2009 data")
#china
unemployment_china <- unemployment_data_2[unemployment_data_2$Country.Code=="CHN",]
auto.arima(unemployment_china$unemp[32:50])
fit <-arima(x=unemployment_china$unemp[32:50], order=c(1,0,0))
preds <-forecast.Arima(fit, h=12)
plot.forecast(preds, xlab = "USA Unemployment forecast using 2000-2009 data")
```


Attempt to find trends in the CPI data.
This data set contains CPI for 265 countries over a period of 56 years.
The variable CPI is time sensitive.
We have used Dickey-fuller test to see if the time series is stationary and checked for p and q values through auto correlation function and partial auto correlation function.

```{r}
CPI_trend<-read.csv("CPI_World_Bank.csv")
#plot(CPI_trend$)
colnames(CPI_trend)
#goal here is time period on x and CPI on y.
trends<-read.csv("Timeseries.csv")
class(trends)
#summary(lm(Time ~ CPI_USA, data=trends))
#plot(trends)
#abline(lsfit(x=trends$Time,y=trends$CPI_USA),col="red")
#install.packages('tseries') 
require(tseries) 
#install.packages('xts')
require(xts)
#converting data into time series data
trends.ts<-ts(as.vector(trends), start = c(1960,1), end = c(2015),frequency = 1)
trends.ts
class(trends.ts)
plot(trends.ts)
#Dickey-fuller test to see if the time series is stationary.
adf.test(trends.ts)
#Checking for p and q values through auto correlation function and partial auto correlation function.
acf(trends.ts)
pacf(trends.ts)
#this is an AR model
#install.packages('forecast')
library('forecast')
auto.arima(trends.ts)
trends.arimausa<- arima(trends.ts, order=c(1,2,1))
trends.arimausa
trend.preds<-forecast.Arima(trends.arimausa,h=12)
plot.forecast(trend.preds)
summary(trend.preds)

#Hypothesis testing for USA CPI data
#install.packages('lmtest')
library('lmtest')
coeftest(trends.arimausa)
#z value is 2.2014, p=0.02771
#another way to obtain p value to verify
(1-pnorm(abs(trends.arimausa$coef)/sqrt(diag(trends.arimausa$var.coef))))*2
```

p value obtained from this analysis is 0.027. Since CPI was an AR series, we will analyze this further on the statistical significance and its effect on the hypothesis.

Provide a bullet-list of the next 5-10 tasks you will perform in analyzing your dataset

1. Compare the time series prediction with the actual data and see how well did the prediction work by checking statistical significance.
2. Improve the time series analysis across all the predictors and see how to convert our dataframe into timeseries for easier analysis.
3. Add more predictors to the dataset and analyze correlation. For example, we could add mental health to see the effects of the crisis on these predictors. We need to check feasibility of obtaining and merging such data into our analysis.
4. Quantify the propagation of the shock to the predictors due to the 2007 crisis using autocorrelation or other techniques.
5. Visualize the change in predictors on a map using geospatial plotting tools.