---
title: "Preliminary Analysis - Effect of Economic Crisis"
author: "Sneha Vasanth, Janani Kumar, Tigh Bradley, Gaurav Gada"
date: "February 21, 2017"
output: html_document
---

Github link: https://github.com/snehavcs/Data-Science-I-Project

The financial crisis in 2007-2009 occurred as a result of risky financial dealings in the real estate market which spilled over into the world economy at large. The resulting economic downturn in the USA, particularly focusing on the economic changes before and after the collapse of Lehman Brothers in September-October 2008, provides an opportunity to compare impact  of the crisis on national ecnomonies. Canada, China, India, and the USA experienced different economic changes during this period, including dramatic changes in unemployment rates.

This study explores what key economic indicators (GDP, CPI, and Unemployment rate) might have looked like in each of these countries if there had been no economic crisis in 2007-2009.  Arima forecasts are utilized to project how ecnomic trends may have continued if the crisis had not occurred.  The forecast models for the 'healthy' economic state will be compared to what  actually occured in these econmies, providing a measure of the relative economic advantages and disadvantages each country experienced during this period.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(dplyr)
library(forecast)
library(lmtest)
```

Cleaning up the data!

```{r load_cpi}
cpi = read.csv("CPI_World_Bank.csv",check.names = FALSE)
#Get country and indicator column names and remove spaces using make.names
country_indicator_names <- make.names(names(cpi[c(1:4)])) 
#country_indicator_names

#Get year column names
year_column_names <- names(cpi[c(5:ncol(cpi))])
#year_column_names

#Reassign the modified column names to actual column names
names(cpi) <- c(country_indicator_names,year_column_names)
names(cpi)

#Gather all year column names into a new column called Year and assign its values to a column called CPI
clean_cpi <- gather(cpi,Year,CPI,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
cpi_data <- clean_cpi[,!colnames(clean_cpi) %in% c("Indicator.Name","Indicator.Code")]
head(clean_cpi)

unemp = read.csv("Unemployment_ILO.csv",check.names = FALSE)
country_indicator_names <- make.names(names(unemp[c(1:4)])) 
#country_indicator_names

#Get year column names
year_column_names <- names(unemp[c(5:ncol(unemp))])
#year_column_names

#Reassign the modified column names to actual column names
names(unemp) <- c(country_indicator_names,year_column_names)
names(unemp)

#Gather all year column names into a new column called Year and assign its values to a column called unemp
clean_unemp <- gather(unemp,Year,Unemployment.Rate,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
unemp_data <- clean_unemp[,!colnames(clean_unemp) %in% c("Indicator.Name","Indicator.Code")]
head(clean_unemp)

gdp = read.csv("gdp.csv",check.names = FALSE)
country_indicator_names <- make.names(names(gdp[c(1:4)])) 
#country_indicator_names

#Get year column names
year_column_names <- names(gdp[c(5:ncol(gdp))])
#year_column_names

#Reassign the modified column names to actual column names
names(gdp) <- c(country_indicator_names,year_column_names)
names(gdp)

#Gather all year column names into a new column called Year and assign its values to a column called gdp
clean_gdp <- gather(gdp,Year,GDP,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
gdp_data <- clean_gdp[,!colnames(clean_gdp) %in% c("Indicator.Name","Indicator.Code")]
head(clean_gdp)

#Reading Manufacturing percentage data.
manufacturing = read.csv("Manufacturing.csv",check.names = FALSE)
country_indicator_names <- make.names(names(manufacturing[c(1:4)])) 
#country_indicator_names

#Get year column names
year_column_names <- names(manufacturing[c(5:ncol(manufacturing))])
#year_column_names

#Reassign the modified column names to actual column names
names(manufacturing) <- c(country_indicator_names,year_column_names)
names(manufacturing)

#Gather all year column names into a new column called Year and assign its values to a column called gdp
clean_manufacturing <- gather(manufacturing,Year,Manufacturing.Rate,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
manufacturing_data <- clean_manufacturing[,!colnames(clean_manufacturing) %in% c("Indicator.Name","Indicator.Code")]
head(clean_manufacturing)

#Reading Manufacturing percentage data.
trade = read.csv("Trade.csv",check.names = FALSE)
country_indicator_names <- make.names(names(trade[c(1:4)])) 
#country_indicator_names

#Get year column names
year_column_names <- names(trade[c(5:ncol(trade))])
#year_column_names

#Reassign the modified column names to actual column names
names(trade) <- c(country_indicator_names,year_column_names)
names(trade)

#Gather all year column names into a new column called Year and assign its values to a column called gdp
clean_trade <- gather(trade,Year,Trade,-Country.Name,-Country.Code,-Indicator.Name,-Indicator.Code)
trade_data <- clean_trade[,!colnames(clean_trade) %in% c("Indicator.Name","Indicator.Code")]
head(clean_trade)
```

```{r}
# Join all data

#gdp_cpi = merge(gdp_data,unemp_data, by=c("Country.Code","Year"), all = T) # NA's match
#manufacturing_data = merge(manufacturing_data,unemp_data,by=c("Country.Code","Year"), all = T) # NA's match
#economy_data = merge(manufacturing_data,gdp_cpi, by=c("Country.Code","Year"), all = T) # NA's match

economy_data <- left_join(gdp_data,cpi_data, by = c("Country.Code","Country.Name","Year"))%>%        left_join(.,manufacturing_data,by=c("Country.Code","Country.Name","Year")) %>% 
left_join(.,unemp_data, by=c("Country.Code","Country.Name","Year")) %>%
left_join(.,trade_data,by=c("Country.Code","Country.Name","Year"))

dim(economy_data)
names(economy_data)
```


```{r}
economy_data = economy_data[c("Country.Name", "Country.Code", "Year", "GDP", "Unemployment.Rate", "CPI","Manufacturing.Rate","Trade")]

# Clean up data
economy_data = economy_data[!(is.na(economy_data$Unemployment.Rate) & is.na(economy_data$CPI) & is.na(economy_data$GDP) & is.na(economy_data$Manufacturing.Rate) & is.na(economy_data$Trade)),]

#Backing up economy data to obtain region level data later
economy_data_full <- economy_data

country_codes = read.csv("country-codes.csv")

country_data = economy_data[economy_data$Country.Code %in% country_codes$ISO3166.1.Alpha.3,]

#rm(list = ls()[grep("economy_data", ls(), invert = T)])

economy_data <- country_data 

# Clean up the working space
rm(list = ls()[grep("economy_data", ls(), invert = T)])
```


## Exploratory Analysis

How many unique observations to you have?

The number of unique observations for CPI, GDP, unemployment, manufacturing and trade respectively are:

```{r}
nrow(economy_data[!is.na(economy_data$CPI),])
nrow(economy_data[!is.na(economy_data$GDP),])
nrow(economy_data[!is.na(economy_data$Unemployment.Rate),])
nrow(economy_data[!is.na(economy_data$Manufacturing.Rate),])
nrow(economy_data[!is.na(economy_data$Trade),])
```

What information/features/characteristics do you have for each observation?

We modify the scope of our dataset from 1991 to 2014, since unemployment data is only available for this range.

```{r}
economy_data <- economy_data[economy_data$Year>=1991 & economy_data$Year<=2014,]
head(economy_data)
```

What are the min/max/mean/median/sd values for each of these features?

```{r summary_stats_cpi}
#Summary values for CPI
summary(economy_data$CPI,na.rm = TRUE)
sd(economy_data$CPI, na.rm = TRUE)

#Summary values for GDP
summary(economy_data$GDP,na.rm = TRUE)
sd(economy_data$GDP, na.rm = TRUE)

#Summary values for Unemployment
summary(economy_data$Unemployment.Rate,na.rm = TRUE)
sd(economy_data$Unemployment.Rate, na.rm = TRUE)

#Summary values for Manufacturing
summary(economy_data$Manufacturing.Rate,na.rm = TRUE)
sd(economy_data$Manufacturing.Rate, na.rm = TRUE)

#Summary values for Trade
summary(economy_data$Trade,na.rm = TRUE)
sd(economy_data$Trade, na.rm = TRUE)

```

We found the averages for CPI, gdp, unemployment, manufacturing and trade for all the countries.
Now, let's manipulate the dataset to include a column to indicate if the time period for the data is before(1991-2006) or after(2007-2014) the economic crisis.


```{r}
economy_data <- economy_data %>% mutate(Time_Period = ifelse(Year < 2007,"Before Crisis","After Crisis"))
head(economy_data)
```

Converting Time_Period to a factor variable and releveling the labels.

```{r}
economy_data$Time_Period <- as.character(economy_data$Time_Period)
economy_data$Time_Period <- factor(economy_data$Time_Period, levels = c("Before Crisis", "After Crisis"))
levels(economy_data$Time_Period)
```


```{r}
ggplot(economy_data, aes(x=Time_Period, y = Unemployment.Rate))+geom_boxplot(na.rm = T)
```

The boxplot confirms the finding the median unemployment is almost the same before and after crisis.
This is an interesting finding since the 2007 economic crisis led to unemployment. We will try to analyze how the unemployment trends varied post recession by drilling down further.


```{r}
ggplot(economy_data, aes(x=Time_Period, y = CPI))+geom_boxplot(na.rm = T)
```

The median CPI has increased increased post the crisis but the variability has got much less over time.

```{r}
ggplot(economy_data, aes(x=Time_Period, y = GDP))+ylim(0,100000000)+geom_boxplot(na.rm = T)
```

The median gdp has decreased post the crisis but the variability has got much less over time.

```{r}
ggplot(economy_data, aes(x=Time_Period, y = Manufacturing.Rate))+geom_boxplot(na.rm = T)
```

The median manufacturing rate has decreased post the crisis.

```{r}
ggplot(economy_data, aes(x=Time_Period, y = Trade))+geom_boxplot(na.rm = T)
```


What is the distribution of the core features (show a histogram)?

Below are the histograms of the distribution of the CPI, GDP, Unemployment, Manufacturing and Trade before and after 2007 economic crisis.

```{r hist_cpi}
ggplot(economy_data, aes(x = CPI)) +
  geom_histogram(fill = "white",
                 color = "black",
                 binwidth = 5,na.rm = T)+facet_grid(.~Time_Period)
```

The distribution of CPI was significantly different before and after crisis. The count of the number of countries with higher inflation rates increased after crisis.

```{r}

ggplot(economy_data, aes(x = GDP)) +
  geom_histogram(fill = "white",
                 color = "black",na.rm = T)+facet_grid(.~Time_Period)

```


```{r}

ggplot(economy_data, aes(x = Unemployment.Rate)) +
  geom_histogram(fill = "white",
                 color = "black",
                 binwidth = 0.5,na.rm = T)+facet_grid(.~Time_Period)
```


```{r}

ggplot(economy_data, aes(x = Manufacturing.Rate)) +
  geom_histogram(fill = "white",
                 color = "black",na.rm = T)+facet_grid(.~Time_Period)

```

```{r}

ggplot(economy_data, aes(x = Trade)) +
  geom_histogram(fill = "white",
                 color = "black",na.rm = T)+facet_grid(.~Time_Period)

```

Data frame to include only 3 regions of reference: South Asia, USA and EU

```{r}
region_data <- economy_data_full %>% filter(Year>=1991 & Year<=2014 & (Country.Code == 'SAS' | Country.Code == 'NAC' | Country.Code == 'EUU'))
region_data <- region_data %>% mutate(Time_Period = ifelse(Year < 2007,"Before Crisis","After Crisis"))
head(region_data)
region_data$Time_Period <- as.character(region_data$Time_Period)
region_data$Time_Period <- factor(region_data$Time_Period, levels = c("Before Crisis", "After Crisis"))
levels(region_data$Time_Period)
```


Let's understand the effect of CPI,GDP and Unemployment further by analyzing trend graph from 1991 to 2014 in USA,Canada,China and India.

Are there obvious trends in the data (over time, across subgroups, etc.), and are the differences statistically significant?

```{r}
unemp_compare <- region_data %>%
  select(Country.Name,Country.Code,Year,Unemployment.Rate,Time_Period) %>%
  filter(Year >= 1991 & Year<= 2014 & (Country.Code =='EUU'| Country.Code == 'SAS' | Country.Code == 'NAC')) %>%
  group_by(Year)

ggplot(unemp_compare,aes(x = Year, y = Unemployment.Rate, col = Country.Code, group = Country.Code)) +geom_line(na.rm = T,lwd = 1)+ theme(axis.text.x=element_text(angle=90, hjust=1))
```

There is a steep increase in the uemployment rates of North America and Europe while South Asia is comparitively stable post the 2008 recession.


```{r}
manufacturing_compare <- region_data %>%
  select(Country.Name,Country.Code,Year,Manufacturing.Rate) %>%
  filter(Year >= 1991 & Year<= 2014 & (Country.Code =='EUU'| Country.Code == 'SAS' | Country.Code == 'NAC')) %>%
  group_by(Year)

ggplot(manufacturing_compare,aes(x = Year, y = Manufacturing.Rate, col = Country.Code, group = Country.Code)) +geom_line(na.rm = T,lwd = 1)+ theme(axis.text.x=element_text(angle=90, hjust=1))
```

#ToDo: Add interpretation of the above

```{r}
trade_compare <- region_data %>%
  select(Country.Name,Country.Code,Year,Trade) %>%
  filter(Year >= 1991 & Year<= 2014 & (Country.Code =='EUU'| Country.Code == 'SAS' | Country.Code == 'NAC')) %>%
  group_by(Year)

ggplot(trade_compare,aes(x = Year, y = Trade, col = Country.Code, group = Country.Code)) +geom_line(na.rm = T,lwd = 1)+ theme(axis.text.x=element_text(angle=90, hjust=1))
```

#ToDo: Add interpretation of the above

What are the other salient aspects of the data (e.g. geospatial factors, text content, etc.)

The following plots are forecasts for each of the selected country's future unemployment rates based on the 1991-2014 data.  This also provides a projection of what the unemployment rates might have looked like in the absence of the data post the financial crisis

```{r}
par(mfrow=c(1,2))
trends.usa <-region_data %>% select(Year,Unemployment.Rate,Country.Code) %>% filter(Country.Code == "NAC")  
trendsts.usa <- ts(as.vector(trends.usa$Unemployment.Rate),start = c(1991,1), end = c(2014), frequency = 1)

trends.usabc <- region_data  %>% select(Year,Unemployment.Rate,Country.Code) %>% filter(Country.Code == "NAC" & Year >=1991 & Year<=2007)
trendsts.usabc <- ts(as.vector(trends.usabc$Unemployment.Rate),start = c(1991,1), end = c(2007), frequency = 1)

plot(trendsts.usa, ylim = c(2,10), xlim = c(2000,2009))


trends.arimausa<- auto.arima(trendsts.usabc)
trends.arimausa
trend.preds<-forecast.Arima(trends.arimausa,h=12)
plot.forecast(trend.preds, ylim = c(2,10), xlim = c(2000,2009))
summary(trend.preds)
```

```{r}
par(mfrow=c(1,2))
trends.usa <-region_data %>% select(Year,Manufacturing.Rate,Country.Code) %>% filter(Country.Code == "NAC")  
trendsts.usa <- ts(as.vector(trends.usa$Manufacturing.Rate),start = c(1991,1), end = c(2014), frequency = 1)

trends.usabc <- region_data  %>% select(Year,Manufacturing.Rate,Country.Code) %>% filter(Country.Code == "NAC" & Year >=1991 & Year<=2007)
trendsts.usabc <- ts(as.vector(trends.usabc$Manufacturing.Rate),start = c(1991,1), end = c(2007), frequency = 1)

plot(trendsts.usa,ylim = c(10,18), xlim = c(2000,2009))


trends.arimausa<- auto.arima(trendsts.usabc)
trends.arimausa
trend.preds<-forecast.Arima(trends.arimausa,h=12)
plot.forecast(trend.preds, ylim = c(10,18), xlim = c(2000,2009))
summary(trend.preds)
```

```{r}
par(mfrow=c(1,2))
trends.usa <-region_data %>% select(Year,Trade,Country.Code) %>% filter(Country.Code == "NAC")  
trendsts.usa <- ts(as.vector(trends.usa$Trade),start = c(1991,1), end = c(2014), frequency = 1)

trends.usabc <- region_data  %>% select(Year,Trade,Country.Code) %>% filter(Country.Code == "NAC" & Year >=1991 & Year<=2007)
trendsts.usabc <- ts(as.vector(trends.usabc$Trade),start = c(1991,1), end = c(2007), frequency = 1)

plot(trendsts.usa, ylim = c(25,38), xlim = c(2000,2009))


trends.arimausa<- auto.arima(trendsts.usabc)
trends.arimausa
trend.preds<-forecast.Arima(trends.arimausa,h=12)
plot.forecast(trend.preds, ylim = c(25,38), xlim = c(2000,2009))
summary(trend.preds)
```


```{r}
par(mfrow=c(1,2))
trends.usa <-region_data %>% select(Year,GDP,Country.Code) %>% filter(Country.Code == "NAC")  
trendsts.usa <- ts(as.vector(trends.usa$GDP),start = c(1991,1), end = c(2014), frequency = 1)

trends.usabc <- region_data  %>% select(Year,GDP,Country.Code) %>% filter(Country.Code == "NAC" & Year >=1991 & Year<=2007)
trendsts.usabc <- ts(as.vector(trends.usabc$GDP),start = c(1991,1), end = c(2007), frequency = 1)

plot(trendsts.usa, xlim = c(2000,2009))


trends.arimausa<- auto.arima(trendsts.usabc)
trends.arimausa
trend.preds<-forecast.Arima(trends.arimausa,h=12)
plot.forecast(trend.preds, xlim = c(2000,2009))
summary(trend.preds)
```



```{r}
par(mfrow=c(3,2))

#Subset data for NAC region from 1991 and 2014 and convert it to time series

Country_Code_Data <- c("NAC","EUU","SAS")

for(i in 1:length(Country_Code_Data))
{
unemp_reg <-region_data %>% select(Year,Unemployment.Rate,Country.Code) %>% filter(Country.Code == Country_Code_Data[i])  
unemp_reg_ts <- ts(as.vector(unemp_reg$Unemployment.Rate),start = c(1991,1), end = c(2014), frequency = 1)

#Get data only until 2007 for NAC region and convert it to time series

unemp_reg_bc <- unemp_reg %>% filter(Year >=1991 & Year<=2007)
unemp_reg_bc_ts <- ts(as.vector(unemp_reg_bc$Unemployment.Rate),start = c(1991,1), end = c(2007), frequency = 1)

#Get data only until 2007 for NAC region and convert it to time series

unemp_reg_ac <- unemp_reg %>% filter(Year >=1991 & Year<=2008)
unemp_reg_ac_ts <- ts(as.vector(unemp_reg_ac$Unemployment.Rate),start = c(1991,1), end = c(2008), frequency = 1)

#Plotting US Trade data with all data

plot(unemp_reg_ts, xlim = c(2000,2010))

#Arima model for US Trade data only until crisis

mod_unemp_bc<- auto.arima(unemp_reg_bc_ts)
unemp_forecast_bc<-forecast.Arima(mod_unemp_bc)
plot.forecast(unemp_forecast_bc, xlim = c(2000,2010))

#Arima model for US Trade data after crisis until 2008

#mod_cpi_ac<- auto.arima(cpi_reg_ac_ts)
#cpi_forecast_ac<-forecast.Arima(mod_cpi_ac)
#plot.forecast(cpi_forecast_ac, xlim = c(2000,2010))

#Hypothesis testing for NAC Trade Arima Model

#Significance test for ARIMA model with data before
#coeftest(mod_unemp_bc)

#Significance test for ARIMA model with data after crisis until 2008
#coeftest(mod_cpi_ac)
}
```


```{r}
par(mfrow=c(3,2))

#Subset data for NAC region from 1991 and 2014 and convert it to time series

Country_Code_Data <- c("NAC","EUU","SAS")

for(i in 1:length(Country_Code_Data))
{
gdp_reg <-region_data %>% select(Year,GDP,Country.Code) %>% filter(Country.Code == Country_Code_Data[i])  
gdp_reg_ts <- ts(as.vector(gdp_reg$GDP),start = c(1991,1), end = c(2014), frequency = 1)

#Get data only until 2007 for NAC region and convert it to time series

gdp_reg_bc <- gdp_reg %>% filter(Year >=1991 & Year<=2007)
gdp_reg_bc_ts <- ts(as.vector(gdp_reg_bc$GDP),start = c(1991,1), end = c(2007), frequency = 1)

#Get data only until 2007 for NAC region and convert it to time series

gdp_reg_ac <- gdp_reg %>% filter(Year >=1991 & Year<=2008)
gdp_reg_ac_ts <- ts(as.vector(gdp_reg_ac$GDP),start = c(1991,1), end = c(2008), frequency = 1)

#Plotting US Trade data with all data


plot(gdp_reg_ts, xlim = c(1995,2010))

#Arima model for US Trade data only until crisis

mod_gdp_bc<- auto.arima(gdp_reg_bc_ts)
gdp_forecast_bc<-forecast.Arima(mod_gdp_bc)
plot.forecast(gdp_forecast_bc, xlim = c(1995,2015))

#Arima model for US Trade data after crisis until 2008

#mod_cpi_ac<- auto.arima(cpi_reg_ac_ts)
#cpi_forecast_ac<-forecast.Arima(mod_cpi_ac)
#plot.forecast(cpi_forecast_ac, xlim = c(2000,2010))

#Hypothesis testing for NAC Trade Arima Model

#Significance test for ARIMA model with data before
#coeftest(mod_gdp_bc)

#Significance test for ARIMA model with data after crisis until 2008
#coeftest(mod_gdp_ac)
}
```



```{r}
par(mfrow=c(3,2))

#Subset data for NAC region from 1991 and 2014 and convert it to time series

Country_Code_Data <- c("NAC","EUU","SAS")

for(i in 1:length(Country_Code_Data))
{
manu_reg <-region_data %>% select(Year,Manufacturing.Rate,Country.Code) %>% filter(Country.Code == Country_Code_Data[i])  
manu_reg_ts <- ts(as.vector(manu_reg$Manufacturing.Rate),start = c(1991,1), end = c(2014), frequency = 1)

#Get data only until 2007 for NAC region and convert it to time series

manu_reg_bc <- manu_reg %>% filter(Year >=1991 & Year<=2007)
manu_reg_bc_ts <- ts(as.vector(manu_reg_bc$Manufacturing.Rate),start = c(1991,1), end = c(2007), frequency = 1)

#Get data only until 2007 for NAC region and convert it to time series

manu_reg_ac <- manu_reg %>% filter(Year >=1991 & Year<=2008)
manu_reg_ac_ts <- ts(as.vector(manu_reg_ac$Manufacturing.Rate),start = c(1991,1), end = c(2008), frequency = 1)

#Plotting US Trade data with all data

plot(manu_reg_ts, xlim = c(2000,2010))

#Arima model for US Trade data only until crisis

mod_manu_bc<- auto.arima(manu_reg_bc_ts)
manu_forecast_bc<-forecast.Arima(mod_manu_bc)
plot.forecast(manu_forecast_bc, xlim = c(2000,2010))

#Arima model for US Trade data after crisis until 2008

#mod_cpi_ac<- auto.arima(cpi_reg_ac_ts)
#cpi_forecast_ac<-forecast.Arima(mod_cpi_ac)
#plot.forecast(cpi_forecast_ac, xlim = c(2000,2010))

#Hypothesis testing for NAC Trade Arima Model

#Significance test for ARIMA model with data before
coeftest(mod_manu_bc)

#Significance test for ARIMA model with data after crisis until 2008
#coeftest(mod_cpi_ac)

}
```


```{r}
par(mfrow=c(3,2))

#Subset data for NAC region from 1991 and 2014 and convert it to time series

Country_Code_Data <- c("NAC","EUU","SAS")

for(i in 1:length(Country_Code_Data))
{
cpi_reg <-region_data %>% select(Year,Trade,Country.Code) %>% filter(Country.Code == Country_Code_Data[i])  
cpi_reg_ts <- ts(as.vector(cpi_reg$Trade),start = c(1991,1), end = c(2014), frequency = 1)

#Get data only until 2007 for NAC region and convert it to time series

cpi_reg_bc <- cpi_reg %>% filter(Year >=1991 & Year<=2007)
cpi_reg_bc_ts <- ts(as.vector(cpi_reg_bc$Trade),start = c(1991,1), end = c(2007), frequency = 1)

#Get data only until 2007 for NAC region and convert it to time series

cpi_reg_ac <- cpi_reg %>% filter(Year >=1991 & Year<=2008)
cpi_reg_ac_ts <- ts(as.vector(cpi_reg_ac$Trade),start = c(1991,1), end = c(2008), frequency = 1)

#Plotting US Trade data with all data

plot(cpi_reg_ts, xlim = c(2000,2010))

#Arima model for US Trade data only until crisis

mod_cpi_bc<- auto.arima(cpi_reg_bc_ts)
cpi_forecast_bc<-forecast.Arima(mod_cpi_bc)
plot.forecast(cpi_forecast_bc, xlim = c(2000,2010))

#Arima model for US Trade data after crisis until 2008

#mod_cpi_ac<- auto.arima(cpi_reg_ac_ts)
#cpi_forecast_ac<-forecast.Arima(mod_cpi_ac)
#plot.forecast(cpi_forecast_ac, xlim = c(2000,2010))

#Hypothesis testing for NAC Trade Arima Model

#Significance test for ARIMA model with data before
coeftest(mod_cpi_bc)

#Significance test for ARIMA model with data after crisis until 2008
#coeftest(mod_cpi_ac)

}
```


```{r}
CPI_trend<-read.csv("CPI_World_Bank.csv")
#plot(CPI_trend$)
colnames(CPI_trend)
#goal here is time period on x and CPI on y.
trends<-read.csv("Timeseries.csv")
class(trends)
#summary(lm(Time ~ CPI_USA, data=trends))
#plot(trends)
#abline(lsfit(x=trends$Time,y=trends$CPI_USA),col="red")
#install.packages('tseries') 
require(tseries) 
#install.packages('xts')
require(xts)
#converting data into time series data
trends.ts<-ts(as.vector(trends), start = c(1960,1), end = c(2015),frequency = 1)
trends.ts
class(trends.ts)
plot(trends.ts)
#Dickey-fuller test to see if the time series is stationary.
adf.test(trends.ts)
#Checking for p and q values through auto correlation function and partial auto correlation function.
acf(trends.ts)
pacf(trends.ts)
#this is an AR model
auto.arima(trends.ts)
trends.arimausa<- arima(trends.ts, order=c(1,2,1))
trends.arimausa
trend.preds<-forecast.Arima(trends.arimausa,h=12)
plot.forecast(trend.preds)
summary(trend.preds)

#Hypothesis testing for USA CPI data
coeftest(trends.arimausa)
#z value is 2.2014, p=0.02771
#another way to obtain p value to verify
(1-pnorm(abs(trends.arimausa$coef)/sqrt(diag(trends.arimausa$var.coef))))*2

trends.resid <- residuals(trends.arimausa)
Box.test(trends.resid, lag = 10, type = "Ljung-Box")
```

Attempt to find trends in the CPI data.
This data set contains CPI for 265 countries over a period of 56 years.
The variable CPI is time sensitive.
We have used Dickey-fuller test to see if the time series is stationary and checked for p and q values through auto correlation function and partial auto correlation function.

p value obtained from this analysis is 0.027. Since CPI was an AR series, we will analyze this further on the statistical significance and its effect on the hypothesis.

Provide a bullet-list of the next 5-10 tasks you will perform in analyzing your dataset

1. Compare the time series prediction with the actual data and see how well did the prediction work by checking statistical significance.
2. Improve the time series analysis across all the predictors and see how to convert our dataframe into timeseries for easier analysis.
3. Add more predictors to the dataset and analyze correlation. For example, we could add mental health to see the effects of the crisis on these predictors. We need to check feasibility of obtaining and merging such data into our analysis.
4. Quantify the propagation of the shock to the predictors due to the 2007 crisis using autocorrelation or other techniques.
5. Visualize the change in predictors on a map using geospatial plotting tools.